\textbf{The statistical analysis of non-linear longitudinal data in biomedical research using generalized additive models}

### 2021-05-06 Manuscripts/Manuscript v5/Manuscript_AM_v5.Rmd
Working on manuscript on the GAM section

The previous section provided the basic theoretical framework to understand the advantages of GAMs to model non-linear data. Although the specific methods of how GAMs model correlation structures is a topic beyond the scope of this work, it suffices to say that GAMs can handle correlation structures beyond compound symmetry. A detailed description on basis functions and correlations can be found in [@hefley2017]. In a practical sense, when a GAM is implemented for longitudinal data, a spline can be added to the model for the _time_ effect to account for the repeated measures over time. Finally, GAMs can handle missing observations that are MAR or MCAR. To demonstrate these points, this section will use simulation to generate data that follows previously reported non-linear trends in the biomedical literature. The simulated data takes as a starting point the reported longitudinal changes in oxygen saturation in subcutaneous that are reported in Figure 3(C) in [vishwanath2009]. An increase in oxygen saturation in the treatment group is observed through time, while a decrease is seen in the Control group. Following the reported trend, 10 normally distributed observations at each time point with a standard deviation of 10% were simulated.

>>>>

The previous section provided the basic theoretical framework to understand the advantages of GAMs to model non-linear data. Although the specific methods of how GAMs model correlation structures is a topic beyond the scope of this paper, it suffices to say that GAMs can handle correlation structures beyond compound symmetry. A detailed description on basis functions and correlations can be found in [@hefley2017]. In a practical sense, when a GAM is implemented for longitudinal data, a spline can be added to the model for the _time_ effect to account for the repeated measures over time. Finally, GAMs can handle missing observations that are MAR or MCAR. To demonstrate these points, this section will use simulation to generate data that follows previously reported non-linear trends in the biomedical literature.

The simulated data is based on the reported longitudinal changes in oxygen saturation ($StO_2$) in subcutaneous tumors under chemotherapy  that are reported in Figure 3(C) in [vishwanath2009]. An increase in oxygen saturation in the treatment group is observed through time, while a decrease is seen in the Control group. Following the reported trend, 10 normally distributed observations at each time point with a standard deviation (SD) of 10% (approximately matching the SD in the figure) are simulated. The "real" and simulated data appear in Figure \@ref(fig:sim-smooth-plot) A,B.

```{r,sim-tumor-response,echo=FALSE,include=FALSE,message=FALSE,warning=FALSE}

dat<-tibble(StO2=c(4,27,3,2,0.5,7,4,50,45,56),
            Day=rep(c(0,2,5,7,10),times=2),
            Group=as.factor(rep(c("Control","Treatment"),each=5))
            )

## plot the mean response
f1<-ggplot(dat, aes(x = Day, y = StO2, color = Group)) +
    geom_line(size=1,show.legend = FALSE)+
    geom_point(show.legend = FALSE)+
  labs(y=expression(paste('Mean StO'[2],'\n (real)')))+
  theme_classic()+
  scale_color_aaas()+
    scale_x_continuous(breaks=c(0,2,5,7,10))

#This function simulates data for the tumor data using default parameters of 10 observations per time point,and Standard deviation (sd) of 5%.
#Because physiologically StO2 cannot go below 0%, data is  generated with a cutoff value of 0.0001 (the "StO2_sim")

simulate_data <- function(dat, n = 10, sd = 5) {
    dat_sim <- dat %>%
        slice(rep(1:n(), each = n)) %>%
        group_by(Group, Day) %>%
        mutate(
               StO2_sim = pmax(rnorm(n, StO2, sd), 0.0001)) %>%
        ungroup()

    return(dat_sim)
}

n <- 10 #number of observations
sd <- 10 #approximate sd from paper
set.seed(1) #set seed for reproducibility
df <- 6
dat_sim <- simulate_data(dat, n, sd)

#plotting simulated data
f2<-ggplot(dat_sim, aes(x = Day, y = StO2_sim, color = Group)) +
    geom_point(show.legend=FALSE)+
    stat_summary(aes(y = StO2_sim,
                     group=Group), 
                 fun=mean, geom="line",size=1,show.legend = FALSE)+
  labs(y=expression(paste('StO'[2],'\n (simulated)')))+
  theme_classic()+
  scale_color_aaas()+
    scale_x_continuous(breaks=c(0,2,5,7,10))

tumor_sim<-f1+f2+plot_layout(guides = 'collect')+
  plot_annotation(tag_levels='A')&
  ylim(c(0,75)) &
  theme(
     text=element_text(size=18),
     legend.title=element_blank(),
     legend.text = element_blank()
     )&
  scale_color_aaas()

For the data in Figure \@ref(fig:sim-smooth-plot),B it is of interest to know the effect that each treatment has in oxygen saturation. In a practical sense, this implies an interaction model where smooths are created for _Group_ and _Day_, respectively. The main thing to consider is that the interaction needs to be properly structured in the model syntax to account for the fact that one of the variables is numeric ( _Day_ ) and that the other is a factor ( _Group_ ). Because the smooths are centered at 0, the factor variable needs to be specified as a parametric term. Using the package `mgcv` The model then becomes:

`m1<-gam(StO2~Group+s(Day,by=Group,k=5), method='REML',data=dat_sim`

This syntax specifies that the object `m1` will store the model, and that the change in $StO_2$ is modeled by independent smooths for  _Group_ and _Day_ (the parenthesis preceded by _s_) using 5 knots. The parametric term _Group_ is added to separate the effect of treatment, and the `method` chosen to select the smoothing parameters is the restricted maximum likelihood (REML) [@wood2017]. When the smooths are plotted over the raw data, it is clear that the model has been able to capture the trend of the change of $StO_2$ for each group across time (Figure \@ref(fig:sim-smooth-plot),C).

```{r,smooth-plot,echo=FALSE,include=FALSE,message=FALSE,warning=FALSE}

#GAM for StO2

m_1 <- gam(StO2_sim ~ Group+s(Day, by = Group, k = 5),method='REML', data  = dat_sim)

#creates a dataframe using the length of the covariates
f_predict <- expand_grid(Group = factor(c("Control", "Treatment")),
                         Day = seq(0, 10, by = 0.1))

#adds the predictions to the grid and creates a confidence interval
f_predict<-f_predict%>%
    mutate(fit = predict(m_1,f_predict,se.fit = TRUE,type='response')$fit,
           se.fit = predict(m_1, f_predict,se.fit = TRUE,type='response')$se.fit)

#plot smooths and confidence interval
f3<-ggplot(data=dat_sim, aes(x=Day, y=StO2_sim, group=Group)) +
    geom_point(aes(color=Group),size=1.5,alpha=0.5)+
  geom_ribbon(aes( x=Day,ymin=(fit - 2*se.fit), 
                   ymax=(fit + 2*se.fit),fill=Group),
              alpha=0.3,
              data=f_predict,
                inherit.aes=FALSE) +
  geom_line(aes(y=fit,
                color=Group),
              size=1,data=f_predict,
              show.legend = FALSE)+
  #facet_wrap(~Group)+
  labs(y=expression(paste('StO'[2],' (simulated)')))+
    scale_x_continuous(breaks=c(0,2,5,7,10))+
      theme_classic()+
      scale_color_aaas()+
  scale_fill_aaas()

sim_smooth_plot<-f3 & theme(
     text=element_text(size=18),
     legend.title=element_blank()
     )
```

(ref:sim-plot-caption) Simulated data and smooths for oxygen saturation in tumors. A: Real data that follows previously reported trends in tumors under chemotherapy (Treatment) or saline (Control) treatment. B: Simulated data from  a normal distribution with standard deviation of 10% with 10 observations per time point. In A and B lines indicate mean oxygen saturation C: Smooths from the GAM model with interaction of Group and Treatment.Lines represent smooths for each group, shaded regions are 95% confidence intervals.

```{r,sim-smooth-plot,fig.width=10, fig.height=7, out.width='75%', fig.align='center',echo=FALSE,message=FALSE, warning=FALSE,fig.cap='(ref:sim-plot-caption)'}

tumor_sim+sim_smooth_plot

```{r,pairwise-comp,echo=FALSE,include=FALSE,message=FALSE,warning=FALSE}
##Pairwise comparisons

# pdat <- expand.grid(Day = seq(0, 10, length = 400),
#                     Group = c('Treatment', 'Control'))
# 
# smooth_diff <- function(model, newdata, f1, f2, alpha = 0.05,
#                         unconditional = FALSE) {
#     xp <- predict(model, newdata = newdata, type = 'lpmatrix')
#     c1 <- grepl(f1, colnames(xp))
#     c2 <- grepl(f2, colnames(xp))
#     #r1 <- newdata[[var]] == f1
#     #r2 <- newdata[[var]] == f2
#     r1 <- with(newdata, Group == f1)
#     r2 <- with(newdata, Group == f2)
#     ## difference rows of xp for data from comparison
#     X <- xp[r1, ] - xp[r2, ]
#     ## zero out cols of X related to splines for other lochs
#     X[, ! (c1 | c2)] <- 0
#     ## zero out the parametric cols
#     X[, !grepl('^s\\(', colnames(xp))] <- 0
#     dif <- X %*% coef(model)
#     se <- sqrt(rowSums((X %*% vcov(model, unconditional = unconditional)) * X))
#     crit <- qt(alpha/2, df.residual(model), lower.tail = FALSE)
#     upr <- dif + (crit * se)
#     lwr <- dif - (crit * se)
#     data.frame(pair = paste(f1, f2, sep = '-'),
#                diff = dif,
#                se = se,
#                upper = upr,
#                lower = lwr)
# }
# 
# comp1<-smooth_diff(mod_int2,pdat,'Control','Treatment')
# 
# comp_StO2 <- cbind(Day = seq(0, 10, length = 400),
#                    rbind(comp1))
# 
# c1<-ggplot(comp_StO2, aes(x = Day, y = diff, group = pair)) +
#     geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.5,fill='#993336') +
#     geom_line(color='black',size=1) +
#     facet_wrap(~ pair) +
#   theme_classic()+
#     labs(x = 'Days', y = expression(paste('Difference in StO'[2] )))+
#   scale_x_continuous(breaks=c(0,2,5,7,10))+
#   theme(
#      text=element_text(size=18),
#      legend.title=element_blank()
#      )
#   
# 
# 
# c1

### 2021-05-06 Manuscripts/Manuscript v5/Manuscript_AM_v5.tex
Working on manuscript on the GAM section

Specifically, the rationale for the chosen linear and quadratic functions is the likelihood that a measured response in two treatment groups is similar in the initial phase of the study, but as treatment progresses a divergence in the trend of the response indicates a difference in the effect of each treatment. In other words, Group 1 can be thought as a ``Control'' group and Group 2 as a ``Treatment'' group. From the mean response per group (linear or quadratic), the variability or ``error'' of individual responses within each group is simulated using a covariance matrix with compound symmetry (constant variance across time). Thus, the response per subject in both the linear and quadratic simulation corresponds to the mean response per group plus the error (Figure \ref{fig:l-q-response} B,D). A more comprehensive exploration of the fit of rm-ANOVA for linear and non-linear longitudinal data is in Figure \ref{fig:linear-cases} and Figure \ref{fig:quadratic-cases} in the Appendix, where simulation with compound symmetry and independent errors (errors generated from a normal distribution that are not constant over time) and the plot of simulated errors, and fitted parameters in presented.

@@

\hypertarget{gams-and-covariance}{%
\subsection{GAMs and covariance}\label{gams-and-covariance}}

Although the specific methods of how GAMs model correlation structures is a topic beyond the scope of this work, it suffices to say that GAMs can handle correlation structures beyond compound symmetry. A detailed description on basis functions and correlations can be found in {[}52{]}. In a practical sense, when a GAM is implemented for longitudinal data, a spline can be added to the model for the \emph{time} effect to account for the repeated measures over time. An example where this is covered is in the Appendix.

>>>>

Specifically, the rationale for the chosen linear and quadratic functions is the likelihood that a measured response in two treatment groups is similar in the initial phase of the study, but as treatment progresses a divergence in the trend of the response indicates an effect due to treatment. In other words, Group 1 can be thought as a ``Control'' group and Group 2 as a ``Treatment'' group. From the mean response per group (linear or quadratic), the variability or ``error'' of individual responses within each group is simulated using a covariance matrix with compound symmetry (constant variance across time). Thus, the response per subject in both the linear and quadratic simulation corresponds to the mean response per group plus the error (Figure \ref{fig:l-q-response} B,D). A more comprehensive exploration of the fit of rm-ANOVA for linear and non-linear longitudinal data is in Figure \ref{fig:linear-cases} and Figure \ref{fig:quadratic-cases} in the Appendix, where simulation with compound symmetry and independent errors (errors generated from a normal distribution that are not constant over time) and the plot of simulated errors, and fitted parameters in presented.

@@

\hypertarget{the-analyisis-of-longitudinal-biomedical-data-using-gams}{%
\subsection{The analyisis of longitudinal biomedical data using GAMs}\label{the-analyisis-of-longitudinal-biomedical-data-using-gams}}

The previous section provided the basic theoretical framework to understand the advantages of GAMs to model non-linear data. Although the specific methods of how GAMs model correlation structures is a topic beyond the scope of this paper, it suffices to say that GAMs can handle correlation structures beyond compound symmetry. A detailed description on basis functions and correlations can be found in {[}52{]}. In a practical sense, when a GAM is implemented for longitudinal data, a spline can be added to the model for the \emph{time} effect to account for the repeated measures over time. Finally, GAMs can handle missing observations that are MAR or MCAR. To demonstrate these points, this section will use simulation to generate data that follows previously reported non-linear trends in the biomedical literature.

The simulated data is based on the reported longitudinal changes in oxygen saturation (\(StO_2\)) in subcutaneous tumors under chemotherapy that are reported in Figure 3(C) in {[}vishwanath2009{]}. An increase in oxygen saturation in the treatment group is observed through time, while a decrease is seen in the Control group. Following the reported trend, 10 normally distributed observations at each time point with a standard deviation (SD) of 10\% (approximately matching the SD in the figure) are simulated. The ``real'' and simulated data appear in Figure \ref{fig:sim-smooth-plot} A,B.

For the data in Figure \ref{fig:sim-smooth-plot},B it is of interest to know the effect that each treatment has in oxygen saturation. In a practical sense, this implies an interaction model where smooths are created for \emph{Group} and \emph{Day}, respectively. The main thing to consider is that the interaction needs to be properly structured in the model syntax to account for the fact that one of the variables is numeric ( \emph{Day} ) and that the other is a factor ( \emph{Group} ). Because the smooths are centered at 0, the factor variable needs to be specified as a parametric term. Using the package \texttt{mgcv} The model then becomes:

\texttt{m1\textless{}-gam(StO2\textasciitilde{}Group+s(Day,by=Group,k=5),\ method=\textquotesingle{}REML\textquotesingle{},data=dat\_sim}

This syntax specifies that the object \texttt{m1} will store the model, and that the change in \(StO_2\) is modeled by independent smooths for \emph{Group} and \emph{Day} (the parenthesis preceded by \emph{s}) using 5 knots. The parametric term \emph{Group} is added to separate the effect of treatment, and the \texttt{method} chosen to select the smoothing parameters is the restricted maximum likelihood (REML) {[}37{]}. When the smooths are plotted over the raw data, it is clear that the model has been able to capture the trend of the change of \(StO_2\) for each group across time (Figure \ref{fig:sim-smooth-plot},C).

{\centering \includegraphics[width=0.75\linewidth,]{Manuscript_AM_v5_files/figure-latex/sim-smooth-plot-1}

\caption{Simulated data and smooths for oxygen saturation in tumors. A: Real data that follows previously reported trends in tumors under chemotherapy (Treatment) or saline (Control) treatment. B: Simulated data from a normal distribution with standard deviation of 10\% with 10 observations per time point. In A and B lines indicate mean oxygen saturation C: Smooths from the GAM model with interaction of Group and Treatment.Lines represent smooths for each group, shaded regions are 95\% confidence intervals.}\label{fig:sim-smooth-plot}
\end{figure}
